Dialect: A Compact Line Notation Based on SMILES

### 2022-03-16 manuscript.tex
Removes section numbers. closes #92
Unifies use of hydrogens attribute. closes #80

\section{Introduction}

@@

\section{Goals, Design, and Tradeoffs}

@@

\section{Data Model}

@@

\begin{table}
\caption{Data Model Primitives.}
\centering
\begin{tabular}{l l l}
    \hline
    Name & Description & Values \\
    \hline
    \ttt{Option<Type>} & Optional Value & \ttt{None}, \ttt{Type} \\
    \ttt{Index} & Unsigned integer & 0{\ldots}2\textsuperscript{32} \\
    \ttt{Ten} & Integer count & 0{\ldots}10 \\
    \ttt{SignedTen} & Integer count & -9{\ldots}10 \\
    \ttt{Thousand} & Integer count & 0{\ldots}1000 \\
    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
    \ttt{HydrogenCount} & Hydrogen count & \ttt{Implicit}, \ttt{Ten} \\
    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\label{table:primitives}
\end{table}

@@

\section{Constitution}

@@

The atomic \ttt{hydrogens} attribute records the number of associated \textit{virtual hydrogens}. A virtual hydrogen is an atom whose presence is recorded, not as a node and edge, but rather as a unit contribution to an integer tally. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node having a \ttt{hydrogens} attribute set to four. The \ttt{hydrogens} attribute, if present, may assume integer values ranging from zero to nine. Only monovalent hydrogens with unspecified isotopic composition are eligible for virtualization.

The two remaining constitutional attributes support \textit{electron counting}. The purpose of electron counting is to ensure that every valence electron, whether bonding or nonbonding, can be associated with either an atom or a bond. Accurate electron counting is crucial for any molecular representation system given their influence on charge, bond order, and mass.

@@

For convenience, the actual model used by Dialect differs slightly from the above description. Neither nodes nor edges carry an explicit electron count attribute. Instead, edges carry an \ttt{order} attribute, and nodes carry a \ttt{charge} attribute. Bond order equals electron count divided by two. Formal charge equals the number of electrons gained or lost due to ionizations events outside of bond formation. Gaining an electron decrements the \ttt{charge} attribute, whereas losing an electron increments it. If no electrons are gained or lost to ionization outside of bond formation, the formal charge equals zero.

@@

From these rules follow some important consequences. For example, bond order must be an integer greater than zero. This follows from the definition of bond formation. The electron count deducted from each atom is $n$, an integer greater than zero. Given bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$. By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order. Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.

@@

\section{Bond Elision}

To promote compact string representations, Dialect supports \textit{bond elision}. An single bond may be elided by marking it as such. An elided bond nevertheless has a bond order of one. A bond may be elided for one of two reasons. First, elided bonds are not written during serialization, allowing for increased information density. The second reason relates to a feature described in the next section.

\section{Delocalization Subgraph}

@@

\section{Implicit Hydrogens and Subvalence}

In addition to virtual hydrogen count, Dialect supports a second form of hydrogen suppression called \textit{implicit hydrogens}. Like a virtual hydrogen, an implicit hydrogen is monovalent, has only default attributes, and is present as an integer tally associated with a particular atom. But unlike a virtual hydrogen, the presence of an implicit hydrogen can only be deduced through computation. Implicit hydrogens are an integral yet invisible component of many molecular graphs.

@@

\section{Computing Implicit Hydrogen Count}

@@

\section{Atom Index}

@@

\section{Conformation}

@@

\section{Configuration}

@@

\section{Syntax}

@@

\subsection{Atom}

@@

\subsection{Bond}

@@

\subsection{Sequence}

@@

\section{Reading Strings}

@@

\section{Pruning}

@@

\section{Writing Strings}

@@

\section{SMILES Compatibility}

@@

\section{Discussion}

@@

\section{Conclusion}

>>>>

\section*{Introduction}

@@

\section*{Goals, Design, and Tradeoffs}

@@

\section*{Data Model}

@@

\begin{table}
\caption{Data Model Primitives.}
\centering
\begin{tabular}{l l l}
    \hline
    Name & Description & Values \\
    \hline
    \ttt{Option<Type>} & Optional Value & \ttt{None}, \ttt{Type} \\
    \ttt{Index} & Unsigned integer & 0..2\textsuperscript{32} \\
    \ttt{Ten} & Integer count & 0..10 \\
    \ttt{SignedTen} & Integer count & -9..10 \\
    \ttt{Thousand} & Integer count & 0..1000 \\
    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
    \ttt{HydrogenCount} & Hydrogen count & \ttt{Implicit}, \ttt{Ten} \\
    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\label{table:primitives}
\end{table}

@@

\section*{Constitution}

@@

The \ttt{hydrogens} attribute encodes an atom's mode of \textit{hydrogen suppression}. Hydrogen suppression replaces hydrogen atoms in a molecular graph with a more compact representation. Despite the removal of its node, a suppressed hydrogen and its bonds to the parent atom are understood to be present. The abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression a popular strategy for increasing information density.

Dialect supports two mutually-exclusive modes of hydrogen suppression. The first is \textit{virtual hydrogen}. A virtual hydrogen is one whose node and connecting bond have been replaced by an integer tally attribute on the parent atom. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four. When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive. The process of transforming one or more hydrogen nodes into a virtual hydrogen is called "virtualization." The opposite process, transforming one or more virtual hydrogens into a node, is called "reification."

The second mode of suppressed hydrogen is \textit{implicit hydrogen}. An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation. The computation will be described in detail in a subsequent section. Subject to certain limitations, an an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}.

The remaining constitutional atomic attribute, \ttt{charge}, enables electron counting. Electron counting associates each electron in a molecular graph with either an atom or a bond. This is a fundamental requirement for the principle of mass conversion. In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order.\cite{dietz:1995}

@@

For convenience, the actual model used by Dialect differs slightly from the above description. Neither nodes nor edges carry an explicit electron count attribute. Instead, edges carry an \ttt{order} attribute, and nodes carry a \ttt{charge} attribute. Bond order equals electron count divided by two. Formal charge equals the number of electrons gained or lost due to ionization events outside of bond formation. Gaining an electron decrements the \ttt{charge} attribute, whereas losing an electron increments it. If no electrons are gained or lost to ionization outside of bond formation, the formal charge equals zero.

@@

From these rules follow some practical consequences. For one, bond order must be an integer greater than zero. This follows from the definition of bond formation. The electron count deducted from each atom is $n$, an integer greater than zero. Given bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$. By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order. Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.

@@

\section*{Bond Elision}

To promote compact string representations, Dialect supports \textit{bond elision}. A bond of order one may be setting its \ttt{elided} attribute to \ttt{true}. As such, an elided bond always has a bond order of one. A bond may be elided for one of two reasons. First, elided bonds are not written during serialization, allowing for increased information density. The second reason relates to a feature described in the next section.

\section*{Delocalization Subgraph}

@@

\section*{Valence and Subvalence}

@@

\section*{Computing Implicit Hydrogen Count}

@@

\section*{Atom Index}

@@

\section*{Conformation}

@@

\section*{Configuration}

@@

\section*{Syntax}

@@

\subsection*{Atom}

@@

\subsection*{Bond}

@@

\subsection*{Sequence}

@@

\section*{Reading Strings}

@@

\section*{Pruning}

@@

\section*{Writing Strings}

@@

\section*{SMILES Compatibility}

@@

\section*{Discussion}

@@

\section*{Conclusion}
